---
title: "Implementing FACE on your own data"
author: "Elizabeth Bersson"
output:
  html_document:
    toc: true
    toc_depth: 2
---

::: {#intro}
This vignette shows how to implement the FACE model described in
[Feature aware covariance estimation, with application to mixtures of chemical
exposures](https://arxiv.org/abs/2504.08220). It
walks through a complete analysis on a synthetic
chemical exposure dataset while highlighting how to
substitute in your own data. 

The code here is drawn from `SYNTHETIC_APPLICATION.R`, which runs the same
analysis end to end as a plain script. Refer to it if you would rather read the
entire workflow in one place.

<h2 id="toc-heading">Table of contents</h2>
:::

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 7, fig.height = 6, fig.align = "center")
```

```{js move-toc, echo = FALSE}
// move the table of contents below the introduction
document.addEventListener("DOMContentLoaded", function() {
  var toc = document.getElementById("TOC"),
      intro = document.getElementById("intro");
  if (toc && intro) intro.parentNode.insertBefore(toc, intro.nextSibling);
});
```



# When to use FACE

FACE estimates a `p x p` covariance matrix from `n` observations based on \textit{feature covariates}, covariates that vary with the outcome index, and describe auxiliary information about the outcome variables themselves. In a chemical exposure data set, for example, feature covariates might include chemical class, assay time, or chemical-molecular properties like boiling point. 


The sampling model is a Bayesian factor model,

$$ y_i = \Lambda \eta_i + \epsilon_i, \qquad \epsilon_i \sim N_p(0, D), \qquad
   \eta_i \sim N_k(0, I), $$

with a prior that regresses the loadings on the features:

$$ \lambda_j \sim N_k(\Gamma^\top x_j, \; \tau^2 d_j \Theta). $$

Here $x_j$ is the vector of features for variable $j$. Variables with similar
features are shrunk toward similar loadings, and so toward similar covariance
structure. If the features carry no signal, $\Gamma$ shrinks to zero and you
recover a standard factor model — so an uninformative feature costs you little.

**You need two objects:**

| object | dimension | meaning |
|--------|-----------|---------|
| `Y` | `n x p` | your data: $n$ observations, $p$ response variables |
| `X` | `p x q` | your features: $q$ features for each of the $p$ response variables |

Ensure your data matrices are correctly aligned! Row `j` of `X` must
describe column `j` of `Y`.

# Setup

```{r load, message = FALSE}
source("helpers.R")        # helper functions
source("functions.R")      # FACE specific functions

library(parallel)          # used within FACE Gibbs sampler
library(dplyr)             # used for plotting
library(RColorBrewer)      # used for plotting
library(readxl)            # used to import synthetic data

set.seed(123)              # set a seed for reproducibility
```

# Step 1: prepare the data matrix `Y`

Our synthetic example mimics the article's motivating study: it consists of exposures to `r 21` chemicals measured by
`r 2` different sampling methods on
`n = 73` participants. This yields `p = 42` response variables for each participant.

```{r data-create-aux, include = FALSE}
chemnames = c("EHDPP","TCEP","TCPP","TDCPP","TPP","2IPPDPP",
              "24DIPPDPP","4tBPDPP","B4TBPP","DEP","DiBP","BBP",
              "DBP","DEHP","DINP","TOTM","BPA","EPB",
              "MPB","PPB","TCS")
chemclass = c(rep("OPE",9), rep("phthalate",7), rep("phenol",5))
method.names = c("WB","Dust")

aux.data = cbind(paste0("V",1:42),rep(chemnames,2),rep(chemclass,2),rep(method.names,each = length(chemnames)))
colnames(aux.data) = c("Variable","Chemical","Chemical Class","Measurement Source")

aux.covs.xlsx = read_excel("VPandHPVforBetsy.xlsx") |>
  select(3, 4)
colnames(aux.covs.xlsx) = c("Vapor.Pressure", "Production.Volume")
aux.covs.xlsx = aux.covs.xlsx |>
  mutate(#prodVol   = as.numeric(as.factor(prodVol)) - 1,
         Vapor.Pressure = Vapor.Pressure * 133.322) |>
  as.matrix()

aux.data = cbind(aux.data,rbind(aux.covs.xlsx,aux.covs.xlsx))

write.csv(aux.data,"data_dictionary.csv", row.names=F)
```

```{r data-create, include = FALSE}
p1 = length(chemnames)      # chemicals
p2 = length(method.names)   # sampling methods
n  = 73                     # participants
p  = p1 * p2                # total number of variables

# true covariance: strong within-wristband correlation, moderate within-dust
Sigma = matrix(.2, ncol = p, nrow = p)
Sigma[1:p1, 1:p1]         = eye(p1) * (1 - .8) + .8
Sigma[(p1+1):p, (p1+1):p] = eye(p1) * (1 - .5) + .5

Y = rmatnorm(matrix(0, ncol = p, nrow = n), V = Sigma)
Y = scale(Y)
colnames(Y) = paste0("V", 1:p)

write.csv(Y,"synethetic_data.csv", row.names=F)
```

```{r response-data}
# import response data
Y = as.matrix(read.csv("synethetic_data.csv"))
p = ncol(Y)
n = nrow(Y)

# preview the data
Y[1:5,1:6]
dim(Y)
```

**For your data.** `Y` is an `n x p` numeric matrix.

* **Center it.** The model has no intercept, so each column must be mean zero.
* **Consider scaling.** `scale()` puts every variable on a common scale, which
  makes the resulting estimate a correlation-scale object and keeps the prior on
  `D` sensible. For exposure concentrations, log-transform before scaling.
* **Column order matters** — fix it now, because `X` must follow it. Here the
  order is chemical-within-method: chemicals 1--21 for wristbands, then
  chemicals 1--21 for dust.

# Step 2: build the feature matrix `X`


Our data has a corresponding data dictionary that we will use to construct an outcome feature matrix `X` which will consist of one row per variable in `Y`. We will use these auxilary variables to construct meta covariates for our analysis.


```{r auxilary-data}
# import data dictionary
data.dictionary = read.csv("data_dictionary.csv")

# print data
head(data.dictionary)
```

## Categorical features

`getCatDesignMat()` turns a length-`p` vector of group labels into a binary
indicator matrix. Chemical class is the natural example. 

```{r cat}
# construct the feature covariate
class.mat = getCatDesignMat(data.dictionary$Chemical.Class)

dim(class.mat)
colnames(class.mat)
```

## Matrix-variate structure

When each observation is naturally a `p1 x p2` matrix flattened into a length-`p`
vector — here chemical-by-method — `getMatDesignMat(p1, p2)` returns indicators
for row membership and column membership. This lets FACE learn a chemical effect
and a method effect separately.

```{r mat}
# extract matrix dimensions from auxilary data
p1 = length(unique(data.dictionary$Chemical))
p2 = length(unique(data.dictionary$Measurement.Source))

# construct the feature covariate
struct.mat = getMatDesignMat(p1, p2)

dim(struct.mat)
colnames(struct.mat)
```

Note the vectorization convention: `getMatDesignMat` assumes `Y`'s columns were
formed by `c()` on the `p1 x p2` matrix, i.e. all `p1` chemicals for method 1,
then all `p1` chemicals for method 2.

## Continuous features

Any numeric property of a variable can be a column of `X`. Here we read in vapor
pressure and an indicator for high production volume, then `rbind` them so each
chemical's properties appear once for each sampling method.

```{r cts, message = FALSE}
# extract the continuous variables we are interested in
cont.covs = data.dictionary[,c("Vapor.Pressure","Production.Volume")]
# convert a ("yes","no") variable to numeric binary
cont.covs$Production.Volume = as.numeric(as.factor(cont.covs$Production.Volume))-1
```

## Put them together

```{r X}
X = cbind(struct.mat,class.mat,cont.covs) |> 
  as.matrix()

dim(X)          # p x q
colnames(X)
```


**If you have no features at all,** pass `X = NA`. FACE then uses an intercept
column, giving a factor model that shrinks all loadings toward a common value.


# Step 3: group the features with `which.cov.group`

FACE puts a generalized ridge prior on the rows of $\Gamma$, so that each *set*
of related feature columns gets its own shrinkage scale. This matters when your
features are on different scales or of different types.

`which.cov.group` is an integer vector of length `q` labeling which group each
**column of `X`** belongs to. Columns sharing a label share a variance
parameter, which is estimated from the data.

```{r groups}
which.cov.group = c(rep(1, p1),                        # chemical indicators
                    rep(2, p2),                        # measurement source indicators
                    rep(3, length(unique(chemclass))), # chemical class
                    4,                                 # production volume
                    5)                                 # vapor pressure

length(which.cov.group) == ncol(X)
```

Rule of thumb: one group per conceptual block of features. Give each standalone
continuous covariate its own group. Omit the argument entirely (leave it `NA`)
to use a common scale for all features.

# Step 4: run the sampler

```{r fit, results = "hide", cache = T}
model.out = FACE_cusp_GS(Y, X,
                        which.cov.group = which.cov.group,
                        k = p,                      # upper bound on # factors
                        S = 2000,                   # total MCMC iterations
                        burnin = 1000,              # number of MCMC burnin iterations
                        my.seed = 1)                # seed for replicability
```


The arguments worth understanding:

| input | description | 
|----------|--------------|
| `k` | An **upper bound** on the number of latent factors  |
| `S`, `burnin`, `thin` | MCMC parameters |  |
| `my.seed` | seed for reproducibilty | 
| `save.all` | what to output: `0` (default) returns thinned post-burn-in draws for key parameters. `1` returns every iteration. `"ICO"` returns only the running mean of the inverse covariance — use if `p` is very large. |

**Key function outputs**:
`cov` are MCMC samples of the posterior covariance matrix and `cov.inv` MCMC samples of the posterior precision matrix, both are stored as vectorized `p x p`
matrices (one iteration per row). 

# Step 5: summarize the posterior

## Point estimate

The paper reports the Bayes estimate under Stein's loss, which is the inverse of
the posterior mean of the precision matrix:

```{r pointest}
face.cov = qr.solve(matrix(colMeans(model.out$cov.inv), ncol = p))
face.cor = cov2cor(face.cov)
```

## Plot it

```{r heatmap}
col.pal = c(brewer.pal(9, "GnBu")[2], "#FFFFFF", brewer.pal(9, "YlOrRd"))
col.lims = c(min(0, min(face.cor)), 1)

image(x = 1:p, y = 1:p, z = face.cor[, p:1],
      col = col.pal, zlim = col.lims,
      xaxt = "n", yaxt = "n", xlab = "", ylab = "",
      main = "Posterior correlation (FACE)")
# block boundary between the two sampling methods
abline(h = p/2 + .5, v = p/2 + .5, col = "black", lwd = 3)
# chemical class boundaries
for (b in cumsum(table(chemclass))[1:2]) {
  abline(v = c(b, p/2 + b) + .5)
  abline(h = p - c(b, p/2 + b) + .5)
}
legend("bottomleft", inset = 0.052, bg = "white",
       legend = round(seq(col.lims[1], col.lims[2],
                          length.out = length(col.pal)), 2)[c(1,3,6,9,11)],
       fill = col.pal[c(1,3,6,9,11)], cex = 1)
```

The two diagonal blocks — strong correlation within wristbands, moderate within
dust — are the structure we simulated. `SYNTHETIC_APPLICATION.R` has a fuller
version of this plot with class labels on both axes.

# Template for your data

```{r template, eval = FALSE}
source("helpers.R")
source("functions.R")
library(parallel)

## 1. Y: n x p, centered (and usually scaled). Fix the column order here.
Y = scale(your.data.matrix)
n = nrow(Y); p = ncol(Y)

## 2. X: p x q, one ROW per column of Y, in the same order.
X = your.covariate.matrix
stopifnot(nrow(X) == p)

## 3. which.cov.group: length q, one label per COLUMN of X, or NA.
which.cov.group = NA

## 4. Fit.
out = FACE_cusp_GS(Y, X,
                   k = p,
                   S = 20000, burnin = 10000, thin = 1,
                   my.seed = 123,
                   alpha = floor(p/3),
                   a.theta = 1/2, b.theta = 1/2,
                   which.cov.group = which.cov.group)

## 5. Summarize.
cov.hat = qr.solve(matrix(colMeans(out$cov.inv), ncol = p))
cor.hat = cov2cor(cov.hat)
```


